DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-11-01 2015CompleteDerivation/completeDerivation.tex
Second pass for prelude to quantum

In principle, we could simply introduce assumption \ref{???} and work our way to quantum mechanics. Unfortunately, given the history and baggage associated with quantum systems, there is a high chance of being misconstrued. The intuitive pictures people have constructed or inherited are varied, may start from different premises and therefore it is easy to misinterpret what one is saying. Therefore we will first ground in classical mechanics the intuition that will be needed later, making the step into full quantum mechanics a gentler, and hopefully less confusing, one.

We will see that, as a first approximation, quantum particles are more keenly described by classical composite states (i.e. distributions) as, unlike classical particles, they are made of a finite amount of material. We will also see that classical distributions already possess qualitative properties that are typically associated with quantum systems and construct a classical analogue of the uncertainty principle.

\subsection{Particles and distributions}

As we saw in the previous sections, classical particles should not be understood as point-like objects of finite mass as this would open a series of problems and inconsistencies. The issues created by such misidentification are only amplified within the context of quantum mechanics. In fact, many issues preventing a satisfying intuitive picture stem from this confusion.

The problem goes along these line. One assumes, implicitly or explicitly, that quantum particles are point-like. But they don't always behave like one, they also behave like a wave, a distribution. During measurements, this distribution coincides experimentally with a probability distribution. So the state of a quantum particle is a probability distribution. Except that it self-interacts. So its a special quantum probability distribution as the quantum particle behaves like its in more than one place at once. We learn this picture and, sometimes begrudgingly, we get used to it.

Here we contend that this picture is not only confusing and unnecessary, it also prevents to understand fully the deep links between classical and quantum mechanics, in the same way that thinking of classical particles as point-like objects prevented to see a more holistic picture in classical mechanics.

The first thing to note is that all physical objects behave like they are in more than one place at once. The earth is both under Milano and New York. Your right index finger is not, hopefully, at the same position of your left eye. In fact, based on what we saw before, any object of finite mass is a distribution over a region, as discontinuous distributions are ruled out. An electron is an object of finite mass. Why should it be any different? In fact, if one claimed an electron to be point-like, he would be claiming an electron behaves in a different way than all other physical objects. Shouldn't \emph{he} have the burden of proof?

Part of the confusion seem to be the idea that, since the electron is an elementary particle, it should have no spatial extent. But that does not follow. Lack of spatial extend and lack of substructure are independent concepts. In fact: we can construct an infinitesimal object that is not elementary. Suppose we have a sphere with dots moving on its surface. All their dynamics can be expressed by their polar and azimuthal angles. If we scale the sphere, the dynamics stays the same. We can shrink the whole sphere to a point: it is now a point-like object that is not elementary, as there is motion on its surface.

Not only point-like object don't have to be elementary, but it's not clear how they can be elementary. How can the whole universe be reduced to  elements that have no idea how many spatial dimensions are there? Mathematically a set of points by itself does not define the dimension of a space: $\mathbbm{R}^n$ is the same order of infinity for any $n$. One needs a topology, which is an additional structure. If everything is just points, where does this come from?

Entertaining the thought that an electron is actually physically distributed in space may seem preposterous, probably because it's not how we are taught to think about it. But it's no more preposterous than to think it point-like, described by a probability distribution that self-interacts.

The first payoff of accepting that a quantum particle is a finite amount of material distributed in space is that this is more consistent with classical mechanics as we have defined it\footnote{It is also more consistent with general relativity.}. The second payoff is that some features of classical mechanics already explain part of the behavior that is usually thought as inherent to quantum mechanics.

@@

Suppose we have a finite amount of classical material distributed over a single d.o.f. $a_m = \int \rho_\mathcal{c} dq \wedge dp$. The Shannon entropy associated with the distribution, which can be though as the amount of information required to identify a particle of the distribution, is finite as well. $I(\rho_\mathcal{c}) = \int \rho_\mathcal{c} \log (\rho_\mathcal{c})  dq \wedge dp = I_0$. TODO: prove in appendix.

As the state evolves into its final state, the amount of material remains the same: $a_m = \int \rho_{\hat{\mathcal{c}}} dq \wedge dp$. The final entropy will also remain the same since the entropy, like the distribution itself, is just another integrable function: $I(\rho_{\hat{\mathcal{c}}}) = I_0$.

In the same way that each particle is constrained to move along states at equal energy, the composite state is constrained to move along distributions at equal entropy. This makes sense: if the motion is deterministic and reversible, the same amount of information is required to identify a particle of material.

We now ask the following question: once we fix the entropy, what is the relationship between the spread of the marginal distributions along $q$ and $p$? Using Lagrange multipliers, we fix the total amount of material $a_m$ and the total entropy $I_0$ and find the distribution that minimizes the the product of the variance $\sigma_q^2 \sigma_p^2 = \int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp$.
\begin{align*}
L = &\int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp \\
&+ \lambda_1(\int \rho dqdp - \rho_{tot})\\
&+ \lambda_2(- \int \rho \ln \rho \, dqdp - H_0)\\
\delta L = &\int \delta \rho [(q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\ &+ \lambda_1 + \lambda_2 \ln \rho + \lambda_2 ] dqdp = 0 \\
\rho = &\frac{\rho_{tot}}{ \sqrt{4 \pi ^2\sigma_q^2\sigma_p^2}} e^{-\frac{(q-\mu_q)^2(p-\mu_p)^2}{4\sigma_q^2\sigma_p^2}} \\
= &\frac{\rho_{tot}}{e^{H_0 - 1}} e^{-\frac{\pi ^2(q-\mu_q)^2(p-\mu_p)^2}{\exp (2(I_0 - 1))}}
\end{align*}
We find that the distribution that minimizes the spread is the product of two independent Gaussians. Recall that in quantum mechanics, the Gaussian wave packet is the state that minimizes uncertainty: this is the classical analogue.

As the distribution evolves, the entropy is conserved and therefore the product $\sigma_q^2 \sigma_p^2$ can never be less than the one given by the Gaussian distribution of the same entropy. We have:
\begin{align*}
\sigma_q\sigma_p \geq \exp (I_0 - 1) / 2 \pi 
\end{align*}
This is strikingly similar to the Heisenberg uncertainty principle, except that the value depends on the initial state. This can be addressed with some extra assumptions.

Suppose the evolution of the distribution is deterministc and reversible but that each particle is undergoing non deterministic motion. Suppose $\rho$ is a distribution at equilibrium of such motion. While each particle moves randomly, the overall distribution $\rho$ stays the same: the non deterministic motion is effectively just reshuffling the particles within the same distribution. The value of the entropy of the distribution is set by the non deterministic motion of the particles: the stronger it is, the greater the entropy.

Suppose now $\rho$ evolves adiabatically. At each instant we can assume $\rho$ to be at equilibrium. At each instant the particles are reshuffled but the amount of entropy of the distribution remains the same. The overall evolution is still Hamiltonian, but the trajectories predicted by the framework are now the average motion of the particles.

If we assumed that all materials were interacting with the same background that caused all particles to exhibit a random walk of the same magnitude, as defined by the information entropy, then the spread of all distributions for all material will satisfy the same inequality. This is gives us the classical uncertainty principle: $\sigma_q\sigma_p \geq \exp (I_\varnothing - 1) / 2 \pi $ where $I_\varnothing$ is the entropy associated with the background.

Assuming such a background is not far fetched. In fact, it serves as a remainder of the issues mentioned in assumption \ref{ass:infinitesimal_reducibility}: as determinism and reversibility at all scales conceptually does not work, there exists a minimum constant interaction to all particles: the system is not perfectly isolated. If the entropy of our distribution is comparable to the entropy introduced by the background, we are likely in the range where classical mechanics is not valid as the random motion of each particle is not negligible.

The generalization to multiple d.o.f. is outside of the scope of this work. Note, though, that this result is similar in spirit to the Gromov non-squeezing theorem and the symplectic camel, which are already generalized to multiple d.o.f.. It is the link with information theory that gives us further physical insight.

It should now be clear how in classical mechanics we can already find some of the ingredients that are typically assumed to be characteristics of quantum mechanics. It has its own flavor of anti-particles, CPT theorem and uncertainty principle. These are general features of an evolution that is deterministic, reversible and adiabatic (i.e. Hamiltonian dynamics). Quantum mechanics, as we'll see, just inherits these traits.

@@

In this section we turn to a different case in which the relationship between the whole material and its parts is different from the one described in section \ref{sec:reducibility}. The system is still infinitesimally decomposable, made of infinitesimal parts as before, but this time it is irreducible. That is: the state and the evolution of the whole does not tell us anything about the state and the evolution of the parts. As the internal dynamics is undetermined, we'll find that each irreducible element is suitably described by the standard framework of quantum mechanics.

>>>>

While in principle, we could simply introduce assumption \ref{ass:irreducibility} and work our way to quantum mechanics, there is a non insignificant chance of being misconstrued. Different people come with different background, to the point that the same words trigger different intuitions and expectations. Therefore we will first ground in classical mechanics the intuition that will be needed later, making the step into quantum mechanics a gentler, and hopefully less confusing, one. It will also help us in showing that classical and quantum mechanics are not, after all, so different.

We will see that, as a first approximation, quantum particles are more keenly described by classical composite states (i.e. a distribution in phase space) as they are made of a finite amount of material. In fact, classical distributions already possess qualitative properties that are typically associated with quantum systems such as a classical analogue of the uncertainty principle.

\subsection{The case against point particles}

As we defined them in previous sections, classical particles should not be understood as point-like objects of finite mass. It was only by assuming they were infinitesimal parts of an invariant density distribution that gave us conjugate pairs of state variables. Hamiltonian mechanics would not have been justified otherwise.

Any finite amount of classical material is described by a distribution. Therefore, to a first approximation, an electron, which is a finite amount of charge/mass, must be described by a distribution as well. This may seem preposterous at first, but we contend that it is no more preposterous than to think it point-like, described by a probability distribution that self-interacts. In fact, the idea of an electron being a distribution of material is already conceptually more compatible with two issues.

The first one is that an electron does not seem to have a well defined position or momentum. If we misidentify the electron with a classical particle, this would indeed be worth of note and puzzlement. But if the electron is a classical distribution, there is not issue. The earth is both under Milano and New York. Your right index finger is not, hopefully, at the same position of your left eye. In fact, based on what we saw before, any object of finite mass is a distribution over a region, and discontinuous distributions are ruled out by \ref{continuous_map}. An electron is an object of finite mass: why should it be any different? In fact, if one claimed an electron to be point-like, he would be claiming an electron behaves in a different way than all other physical objects of finite mass (which are the only ones we experimentally work with). Shouldn't \emph{he} have the burden of proof?

The other issue is particle-wave duality. If the spread of the distribution is narrow enough for the problem at hand, the motion can be described by the position and center of mass, making it behave like a classical particle. If the spread is not negligible (e.g. it encounters obstacles of a smaller scale) then we have to use the full distribution. Therefore, even in this classical approximation, the electron would have this dual nature.

While this picture is quite reasonable, some people may reject it because they believe that a fundamental object must have no size, it must be a point. But that does not follow: lack of spatial extend and lack of substructure are independent concepts.

In fact, we can construct an infinitesimal object that is not elementary. Suppose we have a sphere with dots moving on its surface. All their dynamics can be expressed by their polar and azimuthal angles. If we scale the sphere, the dynamics stays the same. We can shrink the whole sphere to a point: it is now a point-like object that is not elementary, as there is motion on its surface detectable along different angle of incidence.

Not only point-like objects don't have to be elementary, but it's unclear how can they be elementary. How can the whole universe be reduced to  elements that have no idea how many spatial dimensions are there? Mathematically a set of points by itself does not define the dimension of a space: $\mathbbm{R}^n$ is the same order of infinity for any $n$. One needs a topology to differentiate, which is an additional structure. It's the neighborhoods that define the dimensionality. If everything is just points, where does the topology come from?

If we abandon the idea of point-like objects of finite mass, we avoid these problems and get a picture more consistent across the different realms of physics. And we can also see that classical mechanics already has some features that are thought to be inherent to quantum mechanics.

@@

Suppose we have our classical electron described by a finite amount of classical material distributed over a single d.o.f. $\int \rho_\mathcal{c} dq \wedge dp = 1$ where $\rho_\mathcal{c}$ is the normalized distribution. In this context, we will call \emph{fragment} an infinitesimal part of our classical electron. The Shannon entropy associated with the distribution, which can be though as the amount of information required to identify a fragment given the distribution, is finite as well.\footnote{To be precise, Shannon's differential entropy is actually the number of bits needed relative to a uniform distribution of unitary range on all axis. The intuitive meaning, though, is the same.} $I(\rho_\mathcal{c}) = \int \rho_\mathcal{c} \log (\rho_\mathcal{c})  dq \wedge dp = I_0$. TODO: prove in appendix.

As the state evolves into its final state, the amount of material remains the same: $\int \rho_{\hat{\mathcal{c}}} dq \wedge dp = \int \rho_{\mathcal{c}} dq \wedge dp$. The final entropy will also remain the same: $I(\rho_{\hat{\mathcal{c}}}) = I_0$. Mathematically because $\rho_\mathcal{c} \log (\rho_\mathcal{c})$, like the distribution itself, is just another integrable function. Physically, because the amount of information required to identify a fragment given the distribution has to remain the same under deterministic and reversible evolution. Therefore, in the same way that each fragment is constrained to move along states at equal energy, the composite state is constrained to move along distributions at equal entropy.

We now ask the following question: once we fix the entropy, what is the relationship between the spread of the marginal distributions along $q$ and $p$? Using Lagrange multipliers, we fix the total amount of material to $1$ and the total entropy $I_0$ and find the distribution that minimizes the the product of the variance $\sigma_q^2 \sigma_p^2 = \int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp$.
\begin{align*}
L = &\int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp \\
&+ \lambda_1(\int \rho dqdp - 1)\\
&+ \lambda_2(- \int \rho \ln \rho \, dqdp - I_0)\\
\delta L = &\int \delta \rho [(q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\ &+ \lambda_1 + \lambda_2 \ln \rho + \lambda_2 ] dqdp = 0 \\
\rho = &\frac{1}{ \sqrt{4 \pi ^2\sigma_q^2\sigma_p^2}} e^{-\frac{(q-\mu_q)^2(p-\mu_p)^2}{4\sigma_q^2\sigma_p^2}} \\
= &\frac{1}{e^{I_0 - 1}} e^{-\frac{\pi ^2(q-\mu_q)^2(p-\mu_p)^2}{\exp (2(I_0 - 1))}}
\end{align*}
We find that the distribution that minimizes the spread is the product of two independent Gaussians. Recall that in quantum mechanics, the Gaussian wave packet is the state that minimizes uncertainty: this is the classical analogue.

As the distribution evolves, the entropy is conserved and therefore the product $\sigma_q^2 \sigma_p^2$ can never be less than the one given by the Gaussian distribution of the same entropy. We have:
\begin{align*}
\sigma_q\sigma_p \geq \exp (I_0 - 1) / 2 \pi 
\end{align*}
This is strikingly similar to the Heisenberg uncertainty principle, except that the value depends on the initial state $I_0$.

Suppose, though, the evolution of the distribution is deterministc and reversible while each fragment is undergoing non deterministic motion. Suppose $\rho$ is a distribution at equilibrium of such motion. While each fragment moves randomly, the overall distribution $\rho$ stays the same: the non deterministic motion is effectively just reshuffling the fragments within the same distribution. The value of the entropy of the distribution, and therefore its spread, is set by the non deterministic motion of the particles: the stronger it is, the greater the entropy.

Suppose now the classical electron undergoes quasi-static evolution. At each instant we can assume $\rho$ to be at equilibrium. At each instant the fragments are reshuffled but the amount of entropy of the distribution remains the same. The overall evolution is still Hamiltonian, but the trajectories predicted by the framework are now just the average motion of the fragments.

If we assumed that all our classical electrons were interacting with the same background that caused all fragments to exhibit a random walk of the same magnitude, as defined by the information entropy, then the spread of all distributions for all material would satisfy the same inequality. This gives us the classical uncertainty principle: $\sigma_q\sigma_p \geq \exp (I_\varnothing - 1) / 2 \pi $ where $I_\varnothing$ is the entropy associated with the background.

Assuming such a background is not far fetched. In fact, it serves as a remainder of the issues mentioned in assumption \ref{ass:infinitesimal_reducibility}: as determinism and reversibility at all scales conceptually does not work, there exists a minimum constant interaction: the system cannot be perfectly isolated. If the entropy of our distribution is comparable to the entropy introduced by the background, we are likely in the range where classical mechanics is not valid as the random motion of each fragment is not negligible.

The generalization to multiple d.o.f. is outside of the scope of this work. Note, though, that this result is similar in spirit to the Gromov non-squeezing theorem and the symplectic camel, which are already generalized to multiple d.o.f.. It is the link with information theory that gives us further physical insight and is of particular interest given the strong connection between thermodynamics and Shannon's entropy. This is something to be developed in a future work.

At any rate, this discussion shows, once again, how in classical mechanics we can already find some of the ingredients that are typically assumed to be characteristics of quantum mechanics. It has its own flavor of anti-particles, CPT theorem and uncertainty principle. These should really be considered general features of deterministic and reversible evolution of distributions (i.e. Hamiltonian mechanics). Quantum mechanics just inherits these traits.

The discussion also gives us an intuitive physical picture that we will develop further in the next section: a quantum particle is a well determined distribution of not well determined fragments. The distribution as a whole can be studied and evolves deterministically while the fragments do not.

@@

In this section we explore a different relationship between the whole material and its parts from the one described in section \ref{sec:reducibility}. The system is still infinitesimally decomposable, made of infinitesimal parts as before, but this time it is irreducible. That is: the state and the evolution of the whole does not tell us anything about the state and the evolution of the parts. As the internal dynamics is undetermined, we'll find that each irreducible element is suitably described by the standard framework of quantum mechanics.
